---
title: "Sales Data Analysis Using R"
author: "Shoury Jain"
date: "2026-09-17"
output:
  html_document: default
  pdf_document: default
---

# Sales Data Analysis Using R

## Introduction

This project analyzes sales data using R.

The dataset contains information about products, categories,
quantity sold, unit price, sales amount, and regions.

R is used to organize the data, perform statistical analysis,
create graphs, and understand sales patterns.

### Objectives

- Create sales data
- Check and clean the data
- Perform statistical analysis
- Analyze product sales
- Compare regional sales
- Create graphs
- Interpret the results
- Identify important findings
- Give a conclusion

## Project Background

Sales analysis helps businesses understand their sales performance.

By analyzing sales data, we can compare products, quantities,
prices, and different regions.

In this project, R is used to perform basic data analysis
and visualization.

### Scope of the Project

This project focuses on 20 sales records.

The main variables are:

- Date
- Product
- Category
- Quantity
- Unit Price
- Region
- Sales

## 1. Create Sales Data

We create a sales dataset using the `data.frame()` function.

```{r}
sales <- data.frame(
  Date = c(
    "2026-01-05", "2026-01-08", "2026-01-12", "2026-01-18", "2026-01-23",
    "2026-02-03", "2026-02-09", "2026-02-15", "2026-02-21", "2026-02-27",
    "2026-03-04", "2026-03-10", "2026-03-16", "2026-03-22", "2026-03-28",
    "2026-04-05", "2026-04-11", "2026-04-17", "2026-04-23", "2026-04-29"
  ),
  Product = c(
    "Laptop", "Mouse", "Keyboard", "Monitor", "Headphones",
    "Laptop", "Mouse", "Keyboard", "Monitor", "Headphones",
    "Laptop", "Mouse", "Keyboard", "Monitor", "Headphones",
    "Laptop", "Mouse", "Keyboard", "Monitor", "Headphones"
  ),
  Category = c(
    "Electronics", "Accessories", "Accessories", "Electronics", "Accessories",
    "Electronics", "Accessories", "Accessories", "Electronics", "Accessories",
    "Electronics", "Accessories", "Accessories", "Electronics", "Accessories",
    "Electronics", "Accessories", "Accessories", "Electronics", "Accessories"
  ),
  Quantity = c(
    3, 15, 10, 6, 12,
    4, 18, 12, 7, 15,
    5, 20, 14, 8, 18,
    6, 22, 16, 9, 20
  ),
  Unit_Price = c(
    55000, 1200, 2200, 18000, 2500,
    55000, 1200, 2200, 18000, 2500,
    55000, 1200, 2200, 18000, 2500,
    55000, 1200, 2200, 18000, 2500
  ),
  Region = c(
    "North", "West", "South", "East", "West",
    "South", "North", "East", "West", "South",
    "East", "West", "North", "South", "East",
    "West", "South", "East", "North", "West"
  )
)

sales$Sales <- sales$Quantity * sales$Unit_Price

sales
```

### Dataset Description

The dataset contains 20 sales records.

The variables used in the dataset are:

- **Date:** Date of the sale
- **Product:** Name of the product
- **Category:** Product category
- **Quantity:** Number of units sold
- **Unit Price:** Price of one unit
- **Region:** Sales region
- **Sales:** Total sales amount

## 2. Check the Data

We use `summary()` to get a basic overview of the dataset.

```{r}
summary(sales)
```

The summary provides information about the numerical variables
in the dataset.

## 3. Check Dataset Structure

The `str()` function is used to understand the structure
of the dataset.

```{r}
str(sales)
```

This helps us understand the number of observations,
variables, and their data types.

## 4. Check Missing Values and Duplicates

We check whether the dataset contains missing values
or duplicate records.

```{r}
colSums(is.na(sales))
sum(duplicated(sales))
```

The dataset contains no missing values and no duplicate records.

Therefore, the data is ready for analysis.

## 5. Basic Statistical Analysis

We calculate important statistics for quantity and sales.

```{r}
mean(sales$Quantity)
max(sales$Quantity)
min(sales$Quantity)

mean(sales$Sales)
max(sales$Sales)
min(sales$Sales)
```

These values help us understand the average, highest,
and lowest sales performance.

## 6. Total Sales

We calculate the total sales from all records.

```{r}
total_sales <- sum(sales$Sales)

total_sales
```

The total sales represent the overall sales amount
recorded in the dataset.

## 7. Sales by Product

We calculate total sales for each product.

```{r}
product_sales <- aggregate(
  Sales ~ Product,
  data = sales,
  FUN = sum
)

product_sales
```

This table helps compare the sales performance
of different products.

### Product Sales Graph

```{r}
barplot(
  product_sales$Sales,
  names.arg = product_sales$Product,
  main = "Total Sales by Product",
  xlab = "Product",
  ylab = "Total Sales"
)
```

The graph provides a visual comparison of total sales
between different products.

## 8. Quantity Sold by Product

We calculate the total quantity sold for each product.

```{r}
product_quantity <- aggregate(
  Quantity ~ Product,
  data = sales,
  FUN = sum
)

product_quantity
```

### Quantity Sold Graph

```{r}
barplot(
  product_quantity$Quantity,
  names.arg = product_quantity$Product,
  main = "Quantity Sold by Product",
  xlab = "Product",
  ylab = "Quantity Sold"
)
```

The graph shows the number of units sold for each product.

## 9. Sales by Region

We calculate total sales for each region.

```{r}
region_sales <- aggregate(
  Sales ~ Region,
  data = sales,
  FUN = sum
)

region_sales
```

### Regional Sales Graph

```{r}
barplot(
  region_sales$Sales,
  names.arg = region_sales$Region,
  main = "Total Sales by Region",
  xlab = "Region",
  ylab = "Total Sales"
)
```

The graph helps compare sales performance across regions.

## 10. Average Sales by Product

We calculate the average sales for each product.

```{r}
average_product_sales <- aggregate(
  Sales ~ Product,
  data = sales,
  FUN = mean
)

average_product_sales
```

This analysis helps us understand the average sales
generated by different products.

## 11. Quantity vs Sales

A scatter plot is used to understand the relationship
between quantity sold and sales.

```{r}
plot(
  sales$Quantity,
  sales$Sales,
  main = "Quantity Sold vs Sales",
  xlab = "Quantity Sold",
  ylab = "Sales",
  pch = 19
)
```

The graph shows the relationship between the number
of units sold and the sales amount.

## 12. Unit Price vs Sales

We compare unit price with total sales.

```{r}
plot(
  sales$Unit_Price,
  sales$Sales,
  main = "Unit Price vs Sales",
  xlab = "Unit Price",
  ylab = "Sales",
  pch = 19
)
```

The graph helps us understand the relationship between
unit price and sales.

## 13. Correlation Analysis

Correlation helps us understand the relationship
between quantity sold and sales.

```{r}
cor(sales$Quantity, sales$Sales)
```

A positive value indicates a positive relationship
between the two variables.

## 14. Highest Sales Record

We identify the record with the highest sales.

```{r}
sales[which.max(sales$Sales), ]
```

This gives information about the product, quantity,
price, region, and sales amount of the highest sales record.

## 15. Lowest Sales Record

We identify the record with the lowest sales.

```{r}
sales[which.min(sales$Sales), ]
```

This identifies the record with the lowest sales amount.

## 16. Interpretation of Results

The analysis provides several observations.

- Different products generate different sales amounts.
- Sales performance varies between regions.
- Quantity sold has a relationship with total sales.
- Product-wise analysis helps compare products.
- Regional analysis helps compare different regions.
- Graphs make the sales data easier to understand.

The results are based on the sample dataset used in this project.

## 17. Key Findings

The main findings are:

- Total sales can be calculated using `sum()`.
- Product sales can be compared using `aggregate()`.
- Regional sales can be compared using a bar graph.
- Quantity sold can be analyzed using statistical functions.
- Highest and lowest sales records can be identified.
- Graphs provide a visual representation of sales data.
- The dataset has no missing values.
- The dataset has no duplicate records.

## 18. Limitations

This project has some limitations.

- The dataset contains only 20 records.
- Only a limited number of products are included.
- Only four regions are considered.
- Customer information is not included.
- Profit and expenses are not included.
- The dataset is a small sample.

Therefore, the results should be interpreted according
to the available data.

## 19. Future Scope

The project can be improved by using a larger dataset.

Additional information could include:

- Customer details
- Profit
- Discount
- Salesperson
- Monthly sales
- Customer ratings
- Payment method

Advanced statistical methods and machine learning
could also be used for future sales prediction.

## 20. Conclusion

This project analyzed sales data using R.

The project demonstrated how to create a dataset,
check its structure, identify missing values and duplicates,
perform statistical analysis, and create graphs.

Product-wise and region-wise analysis helped us understand
different aspects of sales performance.

Overall, R provides useful tools for data analysis,
visualization, and interpretation of sales data.

## Final Summary

In this project, we learned how to:

- Create a dataset
- Check data structure
- Check missing values
- Check duplicate records
- Calculate statistics
- Calculate total sales
- Analyze products
- Analyze regions
- Create bar graphs
- Create scatter plots
- Calculate correlation
- Interpret results
- Identify findings
- Discuss limitations
- Suggest future improvements